\documentclass[aspectratio=169]{beamer}
%% For 4:3 aspect ratio:
%% \documentclass{beamer}
\usepackage{../git-course}

\title[git-course]{An Introduction to \gh\ and \gs}
\author{Andrew Edwards}
\date{\today}

\begin{document}
%% Needed to remove 'Figure:' from figure captions:
\setbeamertemplate{caption}{\raggedright\insertcaption\par}

\frame[plain]{
\titlepage
}

\section{Definitions}
\frame{\frametitle{Definitions}
  \bi
 \item GitHub -- a website that hosts your repositories so that you can easily
   share code and collaborate with colleagues.
  \item Repository -- essentially a directory containing all your files for a
    project (plus some files that Git uses). Also used to refer to the GitHub
    page for your project.
  \item Git -- a program that allows you to efficiently save ongoing versions of
    your files (`version control') and easily interact with GitHub.
  \ei

Basically, you work on your files in a repository on your computer, use Git on
your computer when you are happy with some changes, and share the files easily on GitHub.
}


\section{Contents}

\frame{\frametitle{Contents}
\bn
\item Creating -- create a new repository on \gh
\item Cloning -- copying it to your local computer
\item Committing -- the crux of working with Git
\en
  \bigskip
I will demonstrate the basics in this video. For each slide I will
demonstrate what to do on the screen (e.g. some commands to type) and then leave
the slide up. You should pause the video and then repeat what was done by
following the instruction on the slide.
\bigskip

There are also a couple of exercises.

\bigskip
The text in the slides is repeated (for easier reference) in the main notes, and
there is also a link there to access these slides.
}

\section{Creating}
\frame{\frametitle{Creating a new repository}
  \bi
    \item Sign into your \gh\ account, click on the
      \emph{Repositories} tab, and press the \emph{New} button.
    \item Give your repository a name. Let's call it \lst{test}.
      \pause
    \item Check \emph{Initialize this repository with a README}.
    \item Leave \emph{`Add .gitignore`} and \emph{`Add a license`}
      set to \emph{None}.
    \item Click \emph{Create repository}.
  \ei

You now have a new repository on the \gh\ website. Next we will \lst{clone} it
onto your computer.
}

\section{Cloning}
\frame{\frametitle{Cloning your new repository}
  \bi
    \item Copy the full URL (web address) of your test repository.
    \item Open the \gs\ and navigated to your \lst{C:/github} directory
      (or whatever you called it when you created it in the setup instructions
      -- it's the place you are going to save all your Git repositories).
    \item run the following command to \emph{clone} your repository:

      ~~\lst{git clone URL}\\
      where \lst{URL} is the url of your newly created repository (paste should work).
  \ei
  ~\\
  You should now have a subdirectory called \lst{github/test} on your computer.\\

  In \gs, change
  to that directory:\\
  ~~\lst{cd test}
  ~\\

  So `clone' is Git speak for copying something from GitHub onto your local
  computer.
  ~\\
  This example has just one file (the README). But the process is the same for a
  repository with multiple files and multiple directories (the structure is
  fully preserved).
}

\frame{\frametitle{Windows only: Storing your credentials}
  When you are using the \gs\ for the \alert{very first time on Windows}, issue
  the following command:\\
  \bigskip
  \lst{git config --global credential.helper wincred}\\
  \bigskip
  This means that you don't have to repeatedly enter your \gh\ password (just do
  it when you are first prompted).\\
}

\section{Committing}
\begingroup
\small
\frame{\frametitle{Creating and committing a new file}
  \bi
    \item Create a new file, \emph{newFile.txt}, in the \lst{github/test}
      directory.
    \pause
    \item Add a line of text at the start of the file and save it.
    \pause
  \item Check the status of the (\lst{test}) repository by typing
    \lst{git status} in the Git shell.
    \pause
  \item It should say that you have an `Untracked file' called \emph{newFile.txt}.
      You want to tell Git to start tracking it, by using the \lst{git add}
      command:\\
      \lst{git add newFile.txt}
   \pause
 \item Type \lst{git status} again. (If on MacOS you may see mention of a
   \emph{.DS\_Store} file -- ignore that for now).
 \item You should see that the file is listed as a `new file` under `Changes to
   be committed`.
    \item Let's now `commit' it:\\
      \lst{git commit -a -m "Add newFile.txt file."}\\
      The commit message should be a useful message saying what the commit
      encapsulates.\\
    \pause
    \item Push the commit to \gh: \lst{git push}
    \pause
    \item Check (refresh) the \gh\ webpage and see your commit and the uploaded file.
  \ei
}
\endgroup

\frame{\frametitle{What just happened?}
  We just used three of the main Git commands:
  \bi
  \item \lst{git add <filename>} -- tell Git to start keeping track of changes
    to this file. You only need to tell Git this once.
  \item \lst{git commit -a -m "Message."} -- committing your changes, which means tell Git
    you are happy with your edits and want to save them. % You are actually saving
    % a snapshot of your entire repository (all the folders and files in your repository).
  \item \lst{git push} -- this sends your commit to the GitHub website.
  \ei

 You always have your files stored \emph{locally} on your computer (as usual),
 even if you don't \lst{add} them or \lst{commit} changes.

 ~\\

 When you push to GitHub then your colleagues can easily \lst{fetch} (retrieve) them.
}

\frame{\frametitle{Keyboard aliases (shortcuts)}
Now, \lst{git commit -a -m "Message."} is a bit much to type, so we have an
alias for it:

~\\

\lst{git com "Message."}

~\\

This is defined in the \emph{.gitconfig} file you installed in the `git-setup`
instructions.

~\\

The \lst{-a} means `commit all changes of files that Git is tracking`, and \lst{-m} is
to include a message. Since we usually want to do both of
these, \lst{git com "Message."} is a useful shortcut. But it is important to
realise it is an alias if searching online for help.

Similarly:

\lst{git s} -- for \lst{git status}

\lst{git p} -- for \lst{git push}

\lst{git d} -- for \lst{git diff}

From now on we will use the aliases.
}

\frame{\frametitle{Edit \emph{Readme.md}}
  Edit the \emph{Readme.md} file. Add some simple comments describing
  the project such as: ``A test repository for learning Git.''\\
  \bigskip
  \pause
  Look over the changes, commit them, and push them to your \gh\
  repository:\\
  \lst{git s}\\
  \lst{git diff} or the alias \lst{git d} -- this gives a simple look at the differences between the last committed
  version and your current version (of all files; only one in this case) \\
  \lst{git com "Initial edit of Readme.md"}\\
  \lst{git p}\\
  ~\\
  Refresh your \gh\ web page and you should see your text
  (the \emph{Readme.md} file is what is shown on the main page
  of your repo).
}

\frame{\frametitle{Exercise 1: create, edit, and commit \emph{simpleText.txt}}
  \bn
    \item Create a text file \lst{simpleText.txt}
      in your local \lst{test} repository. Add a line of text at the start and
      save it.
    \pause
    \item Predict what \lst{git s} will tell you, \emph{then} type it in the
      Git shell to check.\\
    \pause
    \item Add the file to the repository using the git commands:\\
      \lst{git add simpleText.txt}\\
      \lst{git s} -- not necessary but useful to check you understand what is
      changing before you commit\\
      \lst{git com "Adding simpleText.txt"}\\
      \lst{git p}
    \pause
    \item Add some text to \lst{simpleText.txt}, then \lst{git com "Message"} and \lst{git p}.
    \pause
    \item Repeat this a few times to get the hang of it, while intermittently doing \lst{git s} and
      \lst{git d} to understand what's changing.
    \pause
    \item Keep an eye on your commits by refreshing the \gh\ page.
  \en
}

\frame{\frametitle{Adding multiple files at once -- slide 1}
  Often you add multiple files in a new directory. When you run \lst{git s}, you
  will see a large list of \emph{Untracked files}. They can be added at once by simply
  adding the whole directory.
  \bi
    \item Create a new directory to your \lst{test} repository, using your normal method. Call it \lst{new-stuff}.
    \item Add a few new test files to that directory called
      \lst{test1.txt}, \lst{test2.txt}, etc. Put some example text
      in one or more of them if you want.
    \item On the command line, check the status:\\
      \lst{git s}
    \item You will see a listing showing the \lst{new-stuff} directory in
      \emph{Untracked files}.
%    \item To see the actual files to be committed instead of just the
%      directory:\\
%      \lst{git s -u}
    \item To add all the new files in preparation for a commit,
      issue the command:\\
      \lst{git add new-stuff/}
  \ei
  \bigskip
  Continued...
}

\frame{\frametitle{Adding multiple files at once -- slide 2}
  \bi
    \item Check the status of the repository again:
      \lst{git s}
    \item It will now show all files in \emph{Changes to be committed}
    \item Commit the changes:\\
      \lst{git com "Added new-stuff directory."}
    \item Push the changes to \gh:\\
      \lst{git p}
    \item Check your \gh\ webpage and see your commit and that the files
      have been uploaded.
    \item That works no matter how many files are in your \lst{new-stuff}
      directory.
      \ei
  Exercise 2: Repeat the above with more files, to practice creating multiple
  files in a directory and committing your changes.
}

\frame{\frametitle{Adding multiple files at once -- slide 3}
  \bi
    \item To add multiple files with similar names you can use the wildcard
      \lst{*} symbol.
    \item You just added (told Git to keep track of) the new files in your
      \lst{new-stuff/} directory.
    \item If you add more new files to that directory, you will
      have to tell Git to track those also (since they are new -- you haven't
      told Git about them yet).
    \item Say you have 10 new files in \lst{new-stuff/} called \lst{idea1.txt}, \lst{idea2.txt}, ..., \lst{idea10.txt}.
    \item Instead of typing\\
      \lst{git add new-stuff/idea1.txt}\\
      \lst{git add new-stuff/idea2.txt}\\
      etc. (note the \lst{new-stuff/} directory name there)
      you can just use the wildcard \lst{*}:\\
      \lst{git add new-stuff/idea*.txt}\\
      (or even just \lst{git add new-stuff/*.txt}, or \lst{git add new-stuff/*.*}).
    \item No need to do this now, but this is useful to know.
  \ei
}


\frame{\frametitle{The \emph{.gitignore} file}
  But what if you don't want to add all the files that you create?

  ~\\

  Each repository can have a \emph{.gitignore} file, in the root directory
  of the repository.

  Such a file has names of files (such as \lst{my-secret-notes.txt}) or wildcard names (such as
  \lst{*.pdf} or \lst{*.doc}) that will be completely ignored by Git.

  ~\\

  When sharing a repository with others, you want to share your \emph{code} (for
  example, R or Python code) and maybe data, but generally \emph{not} share the output (such as figures
  that the code generates; more on this later). For reproducible research your colleague (or anyone)
  should be able to run your code to generate the results.

  ~\\

  Some programs you run may make temporary files that don't need to be tracked
  by Git, the names of which should also be included in your \emph{.gitignore}.
}

\frame{\frametitle{The \emph{.gitignore} file}
  When sharing code or collaborating you want to keep your
  repository as clean as possible and not clutter it up with files that other
  people don't need.

  ~\\

  So when you run \lst{git s} and see untracked files that you don't want to be
  tracked, add them (or a suitable wildcard expression) to your
  \emph{.gitignore} file so that they are not added inadvertently.

  ~\\

  This will also simplify your workflow (you don't need to keep being reminded
  that you have untracked files).

  ~\\
  If you are on MacOS and you find that directories have a \emph{.DS\_Store} file
  in them, then create (and add and commit) a \emph{.gitignore} file with \emph{.DS\_Store} as a line.
}

\frame{\frametitle{Thoughts/hints regarding commit messages}
  What to write in \lst{git com "Message"}?\\
  Ideally:
  \bi
    \item Want to describe \emph{what} (and sometimes \emph{why}) you did
       something.
    \item The \emph{how} is not needed since that will be explained by the
       actual changes in the code.
    \item Message should be informative for collaborators (including your future self).
  \ei
  {\red Bad}:\\
  ~~\lst{git com "Tweaked function."}\\
  {\blue Good}:\\
  ~~\lst{git com "Allow plot.biomass() to use extra colours."}\\
  ~\\
  A good rule of thumb is to complete the sentence "This commit will ...".
}


\section{Congratulations}
\frame{\frametitle{That's the basics}
  You have now learnt the basics of using Git. By creating a public repository
  on \gh\ you can now release your code to the world!

  ~\\
  You can also choose the \emph{private repository} option when creating a
  repository, so that you can control who can see it.

  ~\\

  Next we will show how to collaborate with colleagues, which is where the
  usefulness of Git will become more apparent.
}

\end{document}
